---
title: "Getting Started with scMiko"
output:
  html_document:
    theme: united
    df_print: kable
  pdf_document: default
date: 'Compiled: `r Sys.Date()`'
---

```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)
```

## Setup

[R](https://www.r-project.org/) version 4.0 or greater is required. We also recommend installing [R Studio](https://www.rstudio.com/).

To install scMiko, run: 

```{r eval = FALSE}
# Enter commands in R (or R studio)
devtools::install_github(repo = "NMikolajewicz/scMiko")
```

```{r}
# load scMiko
library(scMiko)
```


## scMiko within the Seurat framework 

The scMiko package was developed using the [Seurat](https://satijalab.org/seurat/index.html) framework. Consequently, Seurat objects preprocessed using [sctransform](https://github.com/satijalab/sctransform) are expected in scMiko-implemented workflows. 

### Option 1: Preprocess data using Seurat workflow

To preprocess data, users can either refer to the excellent [vignette](https://satijalab.org/seurat/articles/sctransform_vignette.html) offered by Seurat, or follow the adapted workflow shown here.  

```{r seurat object preprocess}
# load Seurat
library(Seurat)

# load data
count_matrix <- readRDS("../data/ps_count_matrix.rds")
meta_data <-  readRDS("../data/ps_meta_data.rds")

# create and preprocess seurat object
so.query <- CreateSeuratObject(counts = count_matrix, meta.data = meta_data)
so.query <- PercentageFeatureSet(so.query, pattern = "^mt-", col.name = "percent.mt")
so.query <- SCTransform(so.query, method = "glmGamPoi", vars.to.regress = "percent.mt", verbose = FALSE)
so.query <- RunPCA(so.query, verbose = FALSE)
so.query <- RunUMAP(so.query, dims = 1:30, verbose = FALSE)
so.query <- FindNeighbors(so.query, dims = 1:30, verbose = FALSE)
so.query <- FindClusters(so.query, verbose = FALSE)

# visualize
cluster.UMAP(so.query) + theme_void() + theme(legend.position = "none")
```


### Option 2: Preprocess data using scPipeline QC & Preprocessing Module.

Another way to preprocess single-cell data is to use our QC & Preprocessing Module implemented in [scPipeline](scPipeline_module_overview.html). Given a gene x cell count matrix, or cell ranger output, the QC & Preprocessing module applies quality-control filters and preprocesses the data. The output from the module is a Seurat object, compatible with all downstream analysis work flows, and an intuitive dashboard report summarizing the data set and results. 

## Sample stratification

Sample-level meta data is often of interest in single cell analyses. To accommodate this information, we have included sample-wise stratification and visualization throughout our `scPipeline` workflows which can be leveraged by including sample-level information in the `Barcodes` field of the Seurat object meta data. This is an optional feature and will not affect the performance of your analysis if omitted.   

## Species support

The `scMiko` package has been optimized for *human* and *murine* samples. 


<details>
  <summary>**Session Info**</summary>
```{r}
sessionInfo()
```
</details>
