---
title: "Getting Started with scPipeline"
output:
  html_document:
    theme: united
    df_print: kable
  pdf_document: default
date: 'Compiled: `r Sys.Date()`'
---

```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)
```

## Setup

[R](https://www.r-project.org/) version 4.0 or greater is required. We also recommend installing [R Studio](https://www.rstudio.com/).

scPipeline is available as a series of RMarkdown stripts and the latest version of scPipeline source code can be found on [scPipeline github repository](https://github.com/NMikolajewicz/scPipeline). To setup and run scPipeline, download the repository locally:

### Option 1: Clone git repository with Bash

``` bash
git clone https://github.com/NMikolajewicz/scPipeline
cd scPipeline
```

### Option 2: Clone git repository with RStudio

In RStudio, navigate to `file` -> `New Project...` -> `Version Control` -> `Git`. You will then be prompted to provide the git repository information:

    Repository URL: https://github.com/NMikolajewicz/scPipeline
    Project Directory Name: scPipeline
    
Press `Create Project` and you're good to go!

## Required Packages

scPipeline is dependent on our [scMiko](https://github.com/NMikolajewicz/scMiko) package. Please find installation instructions [**here**](articles/install.html) prior to proceeding with scPipeline. 

scPipeline is also dependent on [flexdashboards](https://rstudio.github.io/flexdashboard/index.html) to generate interactive dashboards using R Markdown. The [flexdashboard](https://rstudio.github.io/flexdashboard/index.html) package can be installed from CRAN as follows:

```{r, eval = FALSE}
install.packages("flexdashboard")
```

Additionally, each scPipeline module depends on a variety of R packages to run, and a complete list of required packages can be found at the beginning of each module script. 

To install multiple R packages simultaneously, the following command can be used:

```{r, eval = FALSE}
# example
install.packages(c("Seurat", "presto", "dplyr", "tidyr", "stringr", "RColorBrewer", "ggplot2", "DT", "flexdashboard", "ggrepel", "future",  "ggwordcloud", "scales",  "org.Mm.eg.db", "org.Hs.eg.db","parallel", "doParallel", "foreach"))
```

## Running scPipeline Modules

Running scPipeline modules is like running any other Rmarkdown script:

1. **Open** a scPipeline module script (.Rmd) in RStudio.
2. **Specify** analysis parameters
    * See [**Module Overviews**](articles/scPipeline_module_overview.html) for instructions. 
3. **Run** scPipeline Module. Each scPipeline module can be run in one of two ways:
    1. `knit` document (*recommended*)
        * Renders an .HTML report
    2. Run chunk-by-chunk
        * If `print.inline = TRUE`, figures are printing within the R markdown file. 
        * Does not render an .HTML report. 
        * This option permit users to customize their analyses beyond what has been provided by us. 

## scPipeline .HTML reports

Every scPipeline Module generates an HTML report summarizing the analysis results. These can be viewed using any browser. Given their portability (i.e., self-contained), these can be shared easily shared among users. HTML reports are generated using [flexdashboard](https://rstudio.github.io/flexdashboard/index.html).
