---
title: "Getting Started"
output:
  html_document:
    theme: united
    df_print: kable
  pdf_document: default
date: 'Compiled: `r Sys.Date()`'
---

```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)
```

## Setup

[R](https://www.r-project.org/) version 4.0 or greater is required. We also recommend installing [R Studio](https://www.rstudio.com/).

scPipeline is available as a series of RMarkdown stripts and the latest version of scPipeline source code can be found on [scPipeline GitHut repository](https://github.com/NMikolajewicz/scPipeline). To setup and run scPipeline, download the repository locally:

``` bash
git clone https://github.com/NMikolajewicz/scPipeline
cd scPipeline
```

### Basic resources

Each scPipeline module depends on various R packages to run, and these packages are all listed and loaded at the start of each module script. 

```{r}

```

